---
title: "assignment_0, Methods 3, 2026"
date: "2026-09-17"
output: pdf_document
---

```{=html}
<style type="text/css">
  body{
  font-size: 14pt;
}
</style>
```

```{r setup, include=FALSE}
library(tidyverse, fields, lme4)
knitr::opts_chunk$set(echo = TRUE)
```

REMEMBER: In your portfolio, make sure to include code that can reproduce the answers requested in the exercises below (**MAKE A KNITTED VERSION**). If it does not KNIT, it cannot be part of your portfolio

# Assignment 0: Using multilevel modelling to model hierarchical data

In this assignment we will be investigating the *politeness* dataset of Winter and Grawunder (2012) and apply basic methods of multilevel modelling.

## Dataset

The dataset has been shared on GitHub, so make sure that the csv-file is on your current path. Otherwise you can supply the full path.

```{r}
politeness <- read.csv('politeness.csv') # read in data
```

# Exercises and objectives

The objectives of the exercises of this assignment are:\
1) Learning to recognize hierarchical structures within datasets and describing them\
2) Creating simple multilevel models and assessing their fitness\
3) Working both with linear predictors and categorical predictors

REMEMBER: In your report, make sure to include code that can reproduce the answers requested in the exercises below, i.e. make a knitted version\
REMEMBER: This assignment will be part of your final portfolio

--------------------------------------------------- *Iva* ---------------------------------------------------

## Exercise A - describing the dataset and understanding the nature of its variables

1)  Describe the dataset, such that someone who happened upon this dataset could understand the 7 variables and what they contain. This article,<https://www.sciencedirect.com/science/article/pii/S0095447012000666>, might help (see Table 1)
    i.  in this description, categorise the 7 variables into either nominal, ordinal, interval or ratio\
    ii. include number of observations and missing observations\
    iii. what do you think are the response variables, i.e. what could work as *y* in fitting models?

```{r}
str(politeness)

```

```{r}
n_total <- nrow(politeness)
missing_counts <- colSums(is.na(politeness))
missing_counts
```

```{r}
table(politeness$subject)
table(politeness$scenario)
table(politeness$attitude)
table(politeness$gender)
```

```{r}

var_table <- data.frame(
  
  Variable = c("subject",
               "gender",
               "scenario",
               "attitude",
               "total_duration",
               "f0mn",
               "hiss_count"),
  
  Description = c(
    "An identifier for each participant.",
    "The gender of each participant.",
    "The condition type the participant was exposed to",
    "Speech register condition: polite/formal (pol) vs. informal (inf)",
    "How long the utterance lasted.",
    "Mean fundamental frequency of the utterance in Hz; acoustic correlate of pitch.",
    "Number of noisy breath intakes."
  ),
  
  Type = c(
    "nominal",
    "nominal",
    "nominal",
    "nominal",
    "ratio",
    "ratio",
    "ratio"
  ),
  
  N = n_total,
  
  Missing = c(
    missing_counts["subject"],
    missing_counts["gender"],
    missing_counts["scenario"],
    missing_counts["attitude"],
    missing_counts["total_duration"],
    missing_counts["f0mn"],
    missing_counts["hiss_count"]
  ),
  
   "Role in models" = c(
    "grouping variable",
    "predictor",
    "predictor",
    "predictor",
    "response variable (can be a predictor)",
    "response variable (can be a predictor)",
    "response variable (can be a predictor)"
  ),
  check.names = FALSE
)

knitr::kable(var_table,
             caption = "Description of variables in the politeness dataset",
             row.names = FALSE)
```

Note: Answering all sub-questions for exercise 1 below.

The presented dataset is the *politeness* dataset from Winter and Grawunder (2012).
It contains 224 observations (utterances) and includes 16 subjects (9 female, 7 male).

The data is hierarchical: utterances are nested within subjects, and crossed with
7 different scenarios and 2 attitude conditions. See the table above for a summary
of the variables.

The variables *subject*, *gender*, *scenario*, and *attitude* are nominal because
they are labels or categories with no inherent order (e.g., for *scenario* the numbers
1-7 are just labels for different interaction types, and not a continuous scale).

The variables *total_duration*, *f0mn*, and *hiss_count* are ratio variables because
they are continuous measurements with a true 0.

While there are 224 total observations, *f0mn* has 12 missing values, which means
that any model using *f0mn* as the response will effectively use 212 observations.

The primary response variables in this study are *f0mn*, *total_duration*, and
*hiss_count* because these are the measured acoustic outcomes of the speech.
The other variables are predictors or grouping variables that explain the
variation in those responses.

2)  Do some analyses with *scenario* as an independent variable
    i.  fit a multilevel model with *f0mn* as the dependent variable and *scenario* as the independent variable. Model different intercepts for each *subject*

```{r}

#checking how scenario is currently stored as
class(politeness$scenario)
table(politeness$scenario)

```

```{r}

# Model, in which scenario is plotted as numeric
m_num <- lme4::lmer(f0mn ~ scenario + (1 | subject), data = politeness)

m_num
```

ii. now fit a similar model, but this time have _scenario_ be a nominal variable

```{r}
# Model, in which scenario is plotted as nominal

m_nom <- lme4::lmer(f0mn ~ factor(scenario) + (1 | subject), data = politeness)

m_nom
```

iii. judging from the residual variance, (_sigma(model)_), what difference does it make whether you encode _scenario_ as a categorical variable or as an integer variable? Why do you get the difference?

```{r}
# Checking sigma for each model
sigma(m_num)
sigma(m_nom)
```

The residual standard deviation for the numerical model was 32.11, while for the
nominal model it was 30.67. This indicates that the nominal model explains slightly
more variance in f0mn than the numerical model. This may be so because the numerical
model forces a single linear trend across scenarios 1-7. If the true relationship
between *scenario* and *f0mn* is not such straight line, the numerical model will
leave more unexplained variance. The nominal model allows for more flexibility by
estimating a separate intercept for each scenario, therefore reducing the residual
standard deviation.

iv. plot the model matrices, _X_, from the two models, using _image.plot_ from the _fields_ package; describe the two models, and describe the difference between the two kinds of models fitted from these plots.

```{r}
X_num <- model.matrix(m_num)
X_nom <- model.matrix(m_nom)

dim(X_num)
dim(X_nom)

```

```{r}
fields::image.plot(x = 1:dim(X_nom)[2], y = 1:dim(X_nom)[1], t(X_nom),
                   main = "X matrix: scenario as nominal",
                   xlab = "Predictors",
                   ylab = "Observations (n = 212)",
                   legend.lab = "Weight"
                   )



fields::image.plot(x = 1:dim(X_num)[2], y = 1:dim(X_num)[1], t(X_num), 
           main = "X matrix: scenario as numerical",
           xlab = "Predictors",
           ylab = "Observations (n = 212)",
           legend.lab = "Weight",
           xaxt = "n")

axis(1, 1:dim(X_num)[2], c("Intercept", "Slope"))

```



We transposed the model matrices before the visualization to make the structure
clearer. The first model (X_num) treats *scenario* as a numerical variable.
In the plot the y-axis represents the rows (the 212 observations) and the x-axis
represents the columns (variables). The plot shows 2 bands. The solid purple band
is the Intercept (solid because it is always 1). The other band is the *scenario*
variable, which shows a gradient of colors, which represent the numbers 1 to 7.
This means that the numerical model estimates a single linear slope for scenario.

The second model (X_nom) treats *scenario* as a nominal variable. The plot shows
7 bands. The solid band (this time in red) is again the Intercept. The six
remaining bands are dummy variables for scenarios 2-7. They only consist of 0s
and 1s, which indicates that the model estimates a separate intercept for each
scenario rather than a single slope.

**The difference between the models:** The numerical model uses only 1 parameter
for scenario and therefore assumes a straight-line relationship, while the nominal
model uses 7 parameters, which allows each scenario to have its own independent mean.

This explains the residual variance: the nominal model can fit the data more
loosely because it is more flexible, but requires estimating 6 additional parameters.

--------------------------------------------------- *Dori* ---------------------------------------------------

3)  Which variables should be indicated as nominal/categorical?
    i.  use *factor()* to explicitly change these variables into the nominal/categorical category
Nominal/categorical variables are the ones that can be sorted into different groups, but have no inherent ranking/order between them.
```{r}
politeness$attitude <- factor(politeness$attitude)
politeness$scenario <- factor(politeness$scenario)
politeness$gender <- factor(politeness$gender)
```

--------------------------------------------------- *Florin* ---------------------------------------------------

## Exercise B - pooling - linear regression

1)  Across *scenario's*, Do a no-pooling model, a complete-pooling model and a partial-pooling model with *total_duration* as the dependent variable and *f0mn* as the independent variable. Model individual intercepts for each *scenario* for the partial pooling and the no-pooling models

```{r}
df = politeness

# Complete-pooling model
cp_model <- lm(data = df, formula = total_duration ~ f0mn)

# No-pooling model
np_model <- lm(data = df, formula = total_duration ~ f0mn + scenario - 1)

# Partial-pooling model
pp_model <- lmer(data = df, formula = total_duration ~ f0mn + (1 | scenario))
```

i. compare the residual variance between the three models - which results in explaining most variance?

```{r}
  # We will define a function that compares model variance
compare_var <- function(models, complete_comparison = TRUE){
  # complete_comparison is to be set to FALSE when we want a quick answer
  # regarding the least residual variance and not all of the variances
  
  min_var = Inf
  for (model_name in names(models)){
    model <- models[[model_name]]
    var <- summary(model)$sigma^2
    
    # Outputs all residual variances if complete_comparison is TRUE
    if(complete_comparison){
      cat("Unexplained variance for the", model_name, "=", round(var,2), "\n")
    }
    
    # Compares the variances to determine the lowest
    if (var < min_var){
      min_var <- var
      best_model <- model_name
    }
  }
  
  cat("\nThe model that explains the most variance is the ", best_model, ".", sep="")
  
  # This returns only the minimum unexplained variance
  if(!complete_comparison){
    cat("\n\nUnexplained variance for", best_model, "=", round(min_var,2), "\n")
  }
}

# We name the models in a list then call the function
models <- list("complete-pooling model" = cp_model,
               "no-pooling model" = np_model,
               "partial-pooling model" = pp_model)
compare_var(models)
```

ii. for the partial-pooling model, add different intercepts for each _subject_ and compare the residual variance in this model with the remaining three

```{r, warning=FALSE}
pp_model_new <- lmer(data = df,
                 formula = total_duration ~ f0mn + (1 | scenario) + (1 | subject))

models["partial-pooling model with different subjects intercepts"] <- pp_model_new
compare_var(models, complete_comparison = FALSE)
```

  Compared to the previous models, the partial-pooling model that includes random intercepts for both scenarios and subjects accounts for the most variance.

iii. how would you model a no-pooling model that took both _subject_ and _scenario_ into account?

```{r}
np_model_new <- lm(data = df, formula = total_duration ~ f0mn + scenario - 1 + subject - 1)
```

iv. each _subject_ was exposed to each _scenario_; what is the effect of using the second-level effect (1 + subject:scenario) with the first-level effects already modeled?

```{r, warning = FALSE}
pp_model_newest <- lmer(data = df,
                 formula = total_duration ~ f0mn + (1 | subject:scenario))

models["(1+subject:scenario) model"] <- pp_model_newest
compare_var(models, complete_comparison = FALSE)
```

  This model adds random intercepts for every subject x scenario combination and results in the lowest residual variation. However, because there are only 2 data points per subject x scenario combo, it runs a high risk of overfitting, hurting the degree to which it can be generalized to newer data.

--------------------------------------------------- *Dori* ---------------------------------------------------\
2) Plot a model for each subject, ignoring scenarios for now `total_duration ~ f0mn + (1 | subject)` i. create a plot with a subplot for each subject. This plot should contain the observed values for the dependent values and the fitted lines for each subject, ignoring differences in *scenarios* ii. describe what the plots and the model reveal about the relationship between *total_duration* and *f0mn* iii. why do you think the slope is negative? (check the paper this is based on)

```{r}
# removing rows with NAs from the dataframe as we're working with f0mn, where the NAs are
politeness_no_na <- politeness %>% drop_na()
subj_model <- lmer(total_duration ~ f0mn + (1|subject), data = politeness_no_na)
fixef(subj_model)

# adding the predicted y values to a new dataframe
politeness_yhat <- politeness_no_na %>% 
  mutate(pred = fitted(subj_model))

# plotting the data with color coding by subject
ggplot(data = politeness_yhat, aes(x = f0mn, y = total_duration, color = subject)) + 
  geom_point() +
  xlab("Fundamental Frequency (Hz)") +
  ylab("Total Duration (seconds)") +
  geom_line(aes(y = pred))
```

**Short description of what the model does**: We are predicting total duration 
of the acoustic signal by looking at its fundamental frequency; we are letting 
the intercepts vary for each participant, while keeping their slopes constant.

**Relationship between total duration and fundamental frequency**: From fixed 
effects section of output:  A 0.0866 Hz decrease in fundamental frequency can be 
observed per second increase in duration (similar result from plot). Therefore, 
there is a negative correlation between total duration and 
fundamental frequency. Acoustic signals with longer duration generally have 
lower frequencies. 

**Why the slope is negative**: According to the paper, participants spoke on 
average 0.38 ± 0.18 syllables per second slower in the formal condition; 
and since the fundamental frequency was also found to be lower in the 
formal condition, the negative slope makes sense.

--------------------------------------------------- *Alf* ---------------------------------------------------

## Exercise C - pooling - categorical predictors

1) Fit a partial-pooling model with the dependent variable _f0mn_ and the independent variable _attitude_. For second-level effects, include intercepts for _subject_ and _scenario_.

```{r}
partial_model <- lme4::lmer(
  f0mn ~ attitude + (1|subject) + (1|scenario), # including both intercepts
  data = politeness 
  )
```

i. explain what the first-level effects say about the effect of attitude on fundamental frequency, _f0mn_, and explain what the two second level effects say about the variance around those group level effects

**Answer:** 
To know fixed effects we need to look at the coefficients of the model.

```{r}
partial_model_coef <- coef(summary(partial_model)) #We get the coefficients

partial_model_coef #we load the coefficients
```
We can from this data see that attitude does have an impact on the fundamental frequency with a beta value of -14.72. We know the lmer function orders categorical data alphabetically, making the informal attitude be the intercept. We can therefore say that the first-level effects say that a change in attitude from being informal to polite makes fundamental frequency drop by 14.72 Hz.

We now look at the random effects. 

```{r}
print(lme4::VarCorr(partial_model), comp = "Variance")
```
We see that the variance of subjects lay around 4033 and 105 around scenario. So our data varies more between subjects than it does around the test scenario.

ii. also plot the _X_ and _Z_ matrices using _image.plot()_ (for the _Z_, you will have to apply _as.matrix()_ as _image.plot()_ does not handle sparse matrices). Explain what they mean

**Answer:**
We will first get the values using the lme4 library
```{r}
X <- lme4::getME(partial_model, "X")

Z <- lme4::getME(partial_model, "Z")

Z <- as.matrix(Z)

image.plot(x = 1:dim(X)[2], y = 1:dim(X)[1], t(X), xaxt = "n", # Transposing to 
           #make the our design matrix look more traditional
           xlab = "Predictors", 
           ylab = "Observations (n = 212)",
           main = "Visualisation of categorical X matrix",
           legend.lab = "Weight"
           ) 
axis(1, 1:dim(X)[2], c("Intercept", "Slope"))
image.plot(x = 1:dim(Z)[1], y = 1:dim(Z)[2], Z,
           xlab = "Observations (n)", 
           ylab = "Number of random effects (q)",
           main = "Visualisation of categorical Z matrix",
           legend.lab = "Weight"
           ) 
```
Our modelling of the X matrix shows us how our entire first column is made of 1s as intended with our second column made of the attitudes, either being 1 or 0 depending on if it is either a polite attitude or informal attitude. 

The Z matrix holds two effects. It holds effects of both subject and scenario, as we modeled it to have. We can see how each subject, being each "step" of the big diagonal line" holds 1 values and all have a corresponding scenario represented through the diagonal - essentially being all the different scenarios that subject did. 

iii. fit an identical model, except that you remove _scenario_ from the second-level effects. For both models, find the std. errors associated with the group-level effect of attitude - how do they differ and why?

**Answer:**
We write code to find the standard errors to compare.
```{r}
partial_model2 <- lme4::lmer(
  f0mn ~ attitude + (1|subject), # now only including one second-level effect
  data = politeness 
  )
print("Standard error of model with two second-level effects:")
coef(summary(partial_model))[2,2]
print("Standard error of model with one second-level effects:")
coef(summary(partial_model2))[2,2]

```
We see a higher standard error for the model with only one second-level effect. This is potentially because more second-level effects allow for more between-group variation to be explained, leaving less unexplained variance and therefore less standard error. 


2) Making the second-level effects more complicated -  now fit the two models with the formulas: `f0mn ~ attitude * gender + (1 | subject) + (1 | scenario)` and `f0mn ~ attitude * gender + (gender | subject) + (1 | scenario)` 

**Answer:**
We are fitting a more complicated model.

```{r}
complicated_model_1 <- lme4::lmer(f0mn ~ attitude * gender + (1 | subject) + (1 | scenario), data = politeness)

complicated_model_2 <- lme4::lmer(f0mn ~ attitude * gender + (gender | subject) + (1 | scenario), data = politeness)
```

i. Explain what each of the group-level effects and _subject_ and _scenario_ effects mean
**Answer:**
The group-level effects in the first complicated model are quite similar to the less complicated model. They model how the intercept changes according to both subject and the scenario. In our case, the subject effect is grouping the data by the individual participant's data and the scenario is grouping the data by what the scenario, or essentially what trial, the data was taken in. 

The second complicated model has group-level random effect of gender and subject, which means we are modelling if the difference in gender varies from subject to subject. Do note that a subject in this dataset is only allowed one gender, making it a between-subject variable. Therefore, fitting the model to model how the effect of gender changes according to subject - thereby modelling a slope and intercept for each subject according to their different genders. But since there is only one gender per subject the gender slope is the same as just having a baseline, which only leaves the random intercept, and not really modelling any slope.

ii. compare the within-group standard deviation, $\hat\sigma_y$ between the two models - why don't they differ?
**Answer:**
First we will access the group-level effects for each model.
```{r}
print("The first complicated model")
print(sigma(complicated_model_1))
print("The second complicated model")
print(sigma(complicated_model_2))
```
We see here virtually no difference here. This clearly showcases the problem with modelling the random effect slope of genders of individual subject, by showing how it is not able to explain more variance even though we should be adding more complexity. Where the first model has random intercepts for both subjects and scenarios to explain more variation, the second model also models the difference in gender for each subject. But since each subject only has one layer this is essentially the same as only modelling the random intercepts - hence why we see not real difference in standard deviation. 

iii. fit this model `f0mn ~ attitude * gender + (gender | subject) + (gender | scenario)`; why does it give a singular fit? and why would you expect this given the design of the study?

```{r}
singular_fit_model <- lme4::lmer(f0mn ~ attitude * gender + (gender | subject) + (gender | scenario), data = politeness)

lme4::isSingular(singular_fit_model) # checking if model is singular or near singular
```

This model gives a singular fit due to the "dimensions" of the variance-covariance matrix have estimates of exactly zero. This makes sense since our study only allows for one gender per subject, as explained earlier, but doesn't explain all of it since our earlier model was able to handle it. The last push towards our singular fit here might then be caused by the added (gender | scenario) random effect, since our scenario variable only holds 7 and we are now asking the model to compute the scenario-level intercepts, scenario-level slopes relative to gender, although it is possible, and their correlation - it turns into a lot of different parameters to be computed, but we only have 7 different scenarios to do it with. 
