---
title: "assignment_0, Methods 3, 2026"
date: "2026-09-17"
output: pdf_document
---

```{=html}
<style type="text/css">
  body{
  font-size: 14pt;
}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

REMEMBER: In your portfolio, make sure to include code that can reproduce the answers requested in the exercises below (**MAKE A KNITTED VERSION**). If it does not KNIT, it cannot be part of your portfolio

# Assignment 0: Using multilevel modelling to model hierarchical data

In this assignment we will be investigating the *politeness* dataset of Winter and Grawunder (2012) and apply basic methods of multilevel modelling.

## Dataset

The dataset has been shared on GitHub, so make sure that the csv-file is on your current path. Otherwise you can supply the full path.

```{r, eval=FALSE}
politeness <- read.csv('politeness.csv') # read in data
```

# Exercises and objectives

The objectives of the exercises of this assignment are:\
1) Learning to recognize hierarchical structures within datasets and describing them\
2) Creating simple multilevel models and assessing their fitness\
3) Working both with linear predictors and categorical predictors

REMEMBER: In your report, make sure to include code that can reproduce the answers requested in the exercises below, i.e. make a knitted version\
REMEMBER: This assignment will be part of your final portfolio

--------------------------------------------------- *Iva* ---------------------------------------------------

## Exercise A - describing the dataset and understanding the nature of its variables

1)  Describe the dataset, such that someone who happened upon this dataset could understand the 7 variables and what they contain. This article,<https://www.sciencedirect.com/science/article/pii/S0095447012000666>, might help (see Table 1)
    i.  in this description, categorise the 7 variables into either nominal, ordinal, interval or ratio\
    ii. include number of observations and missing observations\
    iii. what do you think are the response variables, i.e. what could work as *y* in fitting models?
2)  Do some analyses with *scenario* as an independent variable
    i.  fit a multilevel model with *f0mn* as the dependent variable and *scenario* as the independent variable. Model different intercepts for each *subject*
    ii. now fit a similar model, but this time have *scenario* be a nominal variable
    iii. judging from the residual variance, (*sigma(model)*), what difference does it make whether you encode *scenario* as a categorical variable or as an integer variable? Why do you get the difference?
    iv. plot the model matrices, *X*, from the two models, using *image.plot* from the *fields* package; describe the two models, and describe the difference between the two kinds of models fitted from these plots

<<<<<<< HEAD
--------------------------------------------------- *Dori* ---------------------------------------------------

3)  Which variables should be indicated as nominal/categorical?

    i.  use *factor()* to explicitly change these variables into the nominal/categorical category
Nominal/categorical variables are the ones that can be sorted into different groups, but have no inherent ranking/order between them.
```{r}
politeness$attitude <- factor(politeness$attitude)
politeness$scenario <- factor(politeness$scenario)
politeness$gender <- factor(politeness$gender)

```


--------------------------------------------------- *Florin* ---------------------------------------------------
=======
3) Which variables should be indicated as nominal/categorical?  
    i. use _factor()_ to explicitly change these variables into the nominal/categorical category

--------------------------------------------------- *Florin* ---------------------------------------------------  

## Exercise B - pooling - linear regression

1)  Across *scenario's*, Do a no-pooling model, a complete-pooling model and a partial-pooling model with *total_duration* as the dependent variable and *f0mn* as the independent variable. Model individual intercepts for each *scenario* for the partial pooling and the no-pooling models
    i.  compare the residual variance between the three models - which results in explaining most variance?
    ii. for the partial-pooling model, add different intercepts for each *subject* and compare the residual variance in this model with the remaining three\
    iii. how would you model a no-pooling model that took both *subject* and *scenario* into account?
    iv. each *subject* was exposed to each *scenario*; what is the effect of using the second-level effect (1 + subject:scenario) with the first-level effects already modeled?

--------------------------------------------------- *Dori* ---------------------------------------------------\
2) Plot a model for each subject, ignoring scenarios for now `total_duration ~ f0mn + (1 | subject)` i. create a plot with a subplot for each subject. This plot should contain the observed values for the dependent values and the fitted lines for each subject, ignoring differences in *scenarios* ii. describe what the plots and the model reveal about the relationship between *total_duration* and *f0mn* iii. why do you think the slope is negative? (check the paper this is based on)

```{r}
library(lme4)
library(tidyverse)

# removing rows with NAs from the dataframe as we're working with f0mn, where the NAs are
politeness_no_na <- politeness %>% drop_na()
subj_model <- lmer(total_duration ~ f0mn + (1|subject), data = politeness_no_na)
fixef(subj_model)

# adding the predicted y values to a new dataframe
politeness_yhat <- politeness_no_na %>% 
  mutate(pred = fitted(subj_model))

# plotting the data with color coding by subject
ggplot(data = politeness_yhat, aes(x = f0mn, y = total_duration, color = subject)) + 
  geom_point() +
  xlab("Fundamental Frequency (Hz)") +
  ylab("Total Duration (seconds)") +
  geom_line(aes(y = pred))
```

**Short description of what the model does**: We are predicting total duration 
of the acoustic signal by looking at its fundamental frequency; we are letting 
the intercepts vary for each participant, while keeping their slopes constant.

**Relationship between total duration and fundamental frequency**: From fixed 
effects section of output:  A 0.0866 Hz decrease in fundamental frequency can be 
observed per second increase in duration (similar result from plot). Therefore, 
there is a negative correlation between total duration and 
fundamental frequency. Acoustic signals with longer duration generally have 
lower frequencies. 

**Why the slope is negative**: According to the paper, participants spoke on 
average 0.38 ± 0.18 syllables per second slower in the formal condition; 
and since the fundamental frequency was also found to be lower in the 
formal condition, the negative slope makes sense.

--------------------------------------------------- *Alf* ---------------------------------------------------

## Exercise C - pooling - categorical predictors

1)  Fit a partial-pooling model with the dependent variable *f0mn* and the independent variable *attitude*. For second-level effects, include intercepts for *subject* and *scenario*
    i.  explain what the first-level effects say about the effect of attitude on fundamental frequency, *f0mn*, and explain what the two second level effects say about the variance around those group level effects
    ii. also plot the *X* and *Z* matrices using *image.plot()* (for the *Z*, you will have to apply *as.matrix()* as *image.plot()* does not handle sparse matrices). Explain what they mean
    iii. fit an identical model, except that you remove *scenario* from the second-level effects. For both models, find the std. errors associated with the group-level effect of attitude - how do they differ and why?
2)  Making the second-level effects more complicated - now fit the two models with the formulas: `f0mn ~ attitude * gender + (1 | subject) + (1 | scenario)` and `f0mn ~ attitude * gender + (gender | subject) + (1 | scenario)`
    i.  Explain what each of the group-level effects and *subject* and *scenario* effects mean
    ii. compare the within-group standard deviation, $\hat\sigma_y$ between the two models - why don't they differ?
    iii. fit this model `f0mn ~ attitude * gender + (gender | subject) + (gender | scenario)`; why does it give a singular fit? and why would you expect this given the design of the study?
